\chapter{Decoding the Crystal Recipe: Predicting Protein Crystallization Conditions via Machine Learning} % Main chapter title

\label{chapter:entry} 
%----------------------------------------------------------------------------------------
%	SECTION 1
%----------------------------------------------------------------------------------------

\section{Introduction}\label{sec:introduction}
Protein crystallization is the process of arranging purified protein molecules into a highly ordered, repeating lattice that forms a crystal. This crystalline state is essential for X-ray crystallography, the most widely used method for determining high-resolution protein structures. When X-rays are diffracted by a protein crystal, the resulting patterns allow reconstruction of the electron density and ultimately the atomic structure of the protein. \\
Crystallization is crucial because structural information provides fundamental insights into protein function and interactions which build the basis of structure-based drug design. Here binding sites derived from the structure enables rational development of small molecules that inhibit or modulate the protein's biological function. \\
Although AlphaFold predictions often align remarkably well with experimentally determined structures, they are not a substitute for them. \textcite{Terwilliger2024} argue that AlphaFold accelerates, but cannot replace, experimental structure determination due to its varying local accuracies and occasional failures at the global structural level. This reinforces that, despite the breakthroughs brought by AlphaFold, further methodological innovation in structure determination and crystallography remains essential.
Beyond crystallography, protein crystals are also used in neutron diffraction, cryo-electron microscopy benchmarking and biophysical studies of stability and folding. \\
Because most proteins do not readily form diffraction-quality crystals, the main bottleneck in structural biology is identifying the crystallization conditions under which a given protein will form suitable crystals \parencite{Mall2025}. Despite advances in structure prediction, identifying suitable crystallization conditions, such as buffer type, pH, salts, and precipitants remains largely empirical and often requires screening hundreds of combinations. This process is both time-consuming and costly, with a high failure rate.

At the same time, extensive data from the Protein Data Bank (PDB) and accurate structural predictions from models like AlphaFold2 have become widely available \parencite{Jumper2021}. These resources offer a unique opportunity to explore whether machine learning models can predict suitable crystallization conditions directly from protein sequence and/or predicted structure.


\section{Related Works}

\subsection{Challenge of X-ray Crystallography}
Crystallizing a protein is often the bottleneck in X-ray crystallography. Only a small fraction (roughly 2–10\%) of proteins produce diffraction-quality crystals, meaning over 90\% of crystallization trials fail \parencite{Mall2025}. This trial-and-error process is costly, where $>$70\% of the total expense in structure determination is spent on attempts not producing crystals of diffraction quality \parencite{Mall2025}. Achieving crystals requires finding the right crystallization conditions (e.g. precipitant chemicals, salts, pH, temperature), but currently these must be determined empirically by screening hundreds or thousands of conditions, which demands a lot of protein \parencite{McPherson2013}. This reality motivates computational approaches to predict either whether a protein is likely to crystallize (crystallization propensity) or even which specific conditions might lead to crystals. The former is a widely researched topic with models such as CrystalP2 \parencite{Kurgan2009} or PPCPred \parencite{Mizianty2011}, the latter however has received little to no attention \parencite{Jin2022}.  While crystallization robots have eased the burden of manual search, discovering optimal conditions still means testing thousands of solutions, wasting protein in the process \parencite{Wilson2014}.
Any such predictive model could significantly reduce experimental screening, saving time and cost.

\subsection{Predicting Protein Crystallization}\label{sec:background:predictingPropensity}

As mentioned predicting crystallization propensity is a well established research area. Early {\bf classical ML} relied on hand‐crafted features such as amino‐acid composition, the proteins isoelectric point, hydrophobicity, disorder, predicted structure and classifiers (SVM, RF, LR, GB). \textcite{Matinyan2024} summarizes multiple tools such as: 
\begin{itemize}
	\item {\it XtalPred}/XtalPred‐RF: feature‐distribution scoring.
	\item {\it TargetCrys}: two‐layer SVM ensemble.
	\item {\it Crysalis} (2016): integrated predictions + mutation suggestions.
	\item {\it BCrystal} (2020): XGBoost + SHAP‐based feature selection.
	\item {\it DCFCrystal} (2021): cascaded RF stages (expression → purification → crystallization), with a membrane‐protein branch.
\end{itemize}
These achieved moderate accuracy (60–75\%, MCC~0.4–0.6) but demanded expert feature engineering.
Deep learning automatically extracts sequence patterns as shown in superior performances presented by:
\begin{itemize}
	\item {\it DeepCrystal} (2019): multi‐scale \gls{cnn} on one‐hot sequences; $\sim83\%$ accuracy, MCC~0.66.
	\item {\it CLPred} (2020): \gls{cnn}+BLSTM; accuracy~85\%, MCC~0.70.
	\item {\it ATTCrys} (2021): adds multi‐head attention; MCC~0.72.
	\item Structure‐infused: {\it SADeepCry} (2022) uses autoencoder+self‐attention on sequence+predicted structure; {\it GCmapCrys} (2023) employs GNNs on AlphaFold2 contact maps.
\end{itemize}

\subsection{Predicting Crystallization pH}
While binary predictions of whether a protein will crystallize are informative, a more demanding and practically relevant objective is the prediction of specific crystallization parameters. Among these, the crystallization pH plays a central role, as it strongly influences protein solubility, surface charge distribution, and intermolecular interactions. Unlike general crystallization screening, pH prediction constitutes a continuous, protein-specific regression problem, and remains particularly challenging due to the complex and often protein-dependent interplay between sequence, structure, and solution chemistry. Although thousands of successful crystallization experiments are available through databases such as the Protein Data Bank, the extent to which protein-derived features can be used to reliably infer optimal crystallization pH remains unclear \parencite{Zhang2022}.

Early attempts to relate protein properties to crystallization pH were already undertaken by \textcite{Kantardjieff2004}, who analyzed 9,596 unique proteins from the \gls{pdb} and reported a strong correlation between the protein’s isoelectric point (\gls{pi}) and the crystallization pH. However, this result is now considered highly controversial, as the reported correlation likely arises from a methodological artifact. Specifically, instead of correlating \gls{pi} directly with pH, the quantity pH–\gls{pi} was used, which introduces an inherit dependency. Subsequent analyses using more recent and curated \gls{pdb} datasets failed to reproduce this finding, showing no meaningful correlation between \gls{pi} and crystallization pH. In particular, \textcite{Kirkwood2015} demonstrated that only weak and inconsistent trends could be observed, concluding that such relationships are insufficiently robust to guide crystallization pH selection. Moreover, the reported effects were limited to buffer pH rather than the final solution pH recorded in the \gls{pdb}, further limiting their practical applicability.

\subsection{Predicting Crystallization Chemicals}
However, the pH of the crystallization solution is not the only variable that is tuned for. Moreover, the optimal chemical cocktail and the respective concentrations are tuned for in traditional experimental screens. Thousands of successful crystallization recipes are known (recorded in databases like the Protein Data Bank), but each protein is unique and may crystallize in different conditions, often unpredictably with no known patterns to predict crystallization conditions \parencite{Zhang2022}. 
\textcite{Abrahams2019} investigated whether protein sequence similarity correlates with crystallization cocktails using \gls{blast}. The similarity between crystallization conditions was quantified using three complementary criteria. First, when applicable, the molecular weight of the \gls{peg} was compared. Second, ammonium sulfate concentrations were analyzed similarly. Finally, a more comprehensive distance metric provided by the \href{https://c6.csiro.au/}{C6 web tool} \parencite{Newman2010} was employed, which additionally accounts for factors such as pH and other components of the crystallization cocktail.
Interestingly, even proteins with high sequence similarity did \textbf{not} necessarily crystallize under similar conditions. This finding highlights that small differences in sequence and/or structure can lead to substantially different optimal crystallization cocktails, emphasizing the inherently multidimensional nature of the crystallization problem. Consequently, \textcite{Abrahams2019} argue that crystallization screens based solely on sequence similarity offer little improvement over standard screening approaches. Since sequence similarity alone carries limited information about optimal crystallization conditions alternative or additional features should be considered when designing predictive models.

\textcite{Liao2025} emphasize the importance of crystal packing and intermolecular packing interfaces in determining crystallization conditions. In their work, they present \gls{mascl}, a framework for simulating crystal packing using AlphaFold combined with symmetrical docking. Crystallization conditions are predicted using a patch-based method that quantifies molecular interface similarity between proteins. For a given target protein, proteins with the most similar physicochemical interface descriptors are used as reference points from which test crystallization conditions are chosen. Meaning no de-novo crystallization conditions are constructed. 

This pipeline of constructing a “crystal fingerprint” for a given protein and comparing it to previously crystallized proteins was evaluated on lysozyme, a common model protein in crystallization studies. In this test case, the proposed AAI-PatchBag approach successfully identified conditions yielding crystals with the desired packing characteristics. However, the use of lyzosome as a model system has been repeatedly criticized \parencite{Chayen2001}. Lysozome benchmarks should be assessed with caution, because it is well known and used for its unusually high crystallizability \parencite{Ghosh2023}. It crystallizes across a broad range of pH values without loss of crystal quality \parencite{Iwai2008}, and also tolerates wide variations in temperature and salt concentration \parencite{Ataka1988}.

Nevertheless, for lysozyme specifically, \textcite{Liao2025} show that similarity in crystal packing information has a stronger influence on predicting successful crystallization conditions than sequence or structural homology alone. It should be noted that the different methods used to construct the fingerprint is extremely costly and might be hard to scale.

In contrast, \textcite{Lee2019} introduced a proof-of-concept deep learning model to map protein sequence to de-novo crystallization conditions. They parsed crystallization records from \gls{pdb} entries and framed the task as a multi-label classification: for a given protein sequence, predict which “crystallization terms” (e.g. specific buffers, salts, precipitants like PEG, etc.) appeared in successful recipes for that protein. Essentially, the model learns associations between sequence features and the types of reagents or techniques that tend to be used. Remarkably, this sequence-to-condition model did show predictive power. A simple 1-layer \gls{cnn} could achieve a weighted F1-score around 0.46 on held-out proteins, substantially better than random guessing in this high-dimensional space. The \gls{cnn} outperformed a fully-connected network, suggesting that local sequence motifs (captured by convolutional filters) were informative for certain crystallization agents. For example, the authors noted that hydrophilic or charged residues in the sequence strongly influenced buffer and salt predictions (likely because they affect the protein’s isoelectric point and solubility). This indicates real biochemical signal: proteins rich in acidic/basic residues might require certain pH buffers or salt conditions, etc., whereas hydrophobic patches might correlate with needing precipitating agents like \glspl{peg} or additives.

It’s important to emphasize that this research is still early-stage. An F1 of 0.45 means the model is far from perfectly pinpointing the exact crystallization recipe, but it is better than trial-and-error alone and demonstrates that sequence patterns can inform what conditions are likely to work. As a future direction, the authors suggested incorporating more variables (e.g. predicting optimal pH and temperature as continuous values, not just class labels) and using the approach to focus screening on a smaller set of candidate conditions. However, after careful assessment of the methodology a crucial error arises. As described, a random function on the files was used to split files into training and test set. This disregards the fact that proteins are crystallized numerous times in the \gls{pdb} and are stored in different files while often being crystallized under the same condition. This presents a strong data leakage with a high probability of the same protein-condition combination being present in both training and evaluation set.