\subsection{Supervised Learning}
As explained by \cite{Mohri2018} in supervised learning, a model is trained using a dataset consisting of input features and corresponding target labels. The goal of the learning algorithm is to estimate a function that maps input variables to output labels, allowing the model to generalize to previously unseen examples.

The training of the model can be formulated as an empirical risk minimization problem. 
Given a training dataset 
\[
\mathcal{D} = \{(x_i, y_i)\}_{i=1}^{N},
\]
where \(x_i \in \mathcal{X}\) denotes an input sample and \(y_i \in \mathcal{Y}\) the corresponding target label, the goal is to learn the optimal model parameters \(\theta\).

This is achieved by minimizing the empirical loss over the training data:

\[
\hat{\theta} = \arg\min_{\theta} \frac{1}{N}\sum_{i=1}^{N} \mathcal{L}(f_\theta(x_i), y_i).
\]

Here, \(f_\theta\) denotes a parameterized model with parameters \(\theta\). The function \(\mathcal{L}(\cdot,\cdot)\) is a loss function that measures the discrepancy between the model prediction \(f_\theta(x_i)\) and the true target value \(y_i\). Common examples include the cross-entropy loss for classification tasks and the mean squared error for regression tasks.

The term inside the summation computes the loss for a single training example, while the average over all \(N\) samples represents the empirical risk. The operator \(\arg\min_{\theta}\) denotes the parameter values that minimize this empirical risk. The resulting parameter estimate \(\hat{\theta}\) therefore corresponds to the model that best fits the training data according to the chosen loss function.

For the task of finding optimal crystallization conditions, the input features are derived from the protein sequence. Labels are represented by crystallization condition, where 0 and 1 are used to represent no-/crystal. 

Supervised learning approaches rely heavily on the availability of high-quality labeled datasets. In practice, obtaining reliable negative labels for crystallization experiments can be difficult, which introduces challenges for standard supervised learning formulations. Non-working conditions are often disregarded which makes supervised learning the wrong approach, since a model can only learn if there is variance in the dataset. This is due to label bias which occurs if the model learns more about the distribution of the label than the features in the input. For example, if only positive labels are present (as the case for the \gls{pdb} dataset) the model could achieve perfect performance by predicting that every protein-condition combination results in a crystal, which is far from the truth. In these cases other learning paradigms have to be used. 

\subsection{Contrastive Learning} \label{sec:contrastive_learning}

Contrastive learning has emerged as a powerful paradigm for representation learning, particularly in self-supervised settings where labeled data is scarce. Instead of learning from explicit labels, contrastive methods learn representations by comparing samples and enforcing similarity relationships in an embedding space. As seen in later sections contrastive learning can be especially useful if the labels are insufficient. The core idea is to bring representations of similar samples closer while pushing dissimilar samples apart. This section follows the unified framework explored by \citet{LeKhac2020}.

Let $x \in \mathcal{X}$ denote an input sample. Contrastive learning defines two types of pairs:
\begin{itemize}
    \item \textbf{Positive pairs} $(x, x^+)$: semantically similar samples
    \item \textbf{Negative pairs} $(x, x^-)$: dissimilar samples
\end{itemize}

The goal is to learn a representation function $f_\theta: \mathcal{X} \rightarrow \mathbb{R}^d$ such that:
\begin{equation}
\text{sim}(f_\theta(x), f_\theta(x^+)) \gg \text{sim}(f_\theta(x), f_\theta(x^-))
\end{equation}

This is typically achieved by optimizing a contrastive loss over a batch of such pairs. The general framework consists of four key components \citep{LeKhac2020}:

\paragraph{(1) Data Augmentation and Pair Construction}
Given an input $x$, two stochastic transformations $t, t' \sim \mathcal{T}$ generate two correlated views:
\begin{equation}
x_q = t(x), \quad x_k = t'(x)
\end{equation}
These form a positive pair, while samples from other inputs act as negatives.

\paragraph{(2) Encoder}
A neural network encoder $e(\cdot)$ maps inputs to representations:
\begin{equation}
v = e(x), \quad v \in \mathbb{R}^d
\end{equation}
The encoder is trained to extract meaningful features.

\paragraph{(3) Projection Head}
A projection function $h(\cdot)$ maps representations to a latent space where contrastive loss is applied:
\begin{equation}
z = h(v), \quad z \in \mathbb{R}^{d'}
\end{equation}
This separation allows the encoder to retain general features while the projection head specializes for the contrastive objective.

\paragraph{(4) Contrastive Loss}
The learning signal is provided by a loss function that enforces similarity constraints between embeddings.

\paragraph{Training Objective}

A widely used loss is the InfoNCE loss \citep{LeKhac2020}, defined for a positive pair $(z_i, z_i^+)$ among $K$ negative samples as:
\begin{equation}
\mathcal{L}_i = - \log \frac{\exp(\text{sim}(z_i, z_i^+)/\tau)}
{\sum_{j=0}^{K} \exp(\text{sim}(z_i, z_j)/\tau)}
\end{equation}

where:
\begin{itemize}
    \item $\text{sim}(a,b) = a^\top b$ or cosine similarity
    \item $\tau$ is a temperature parameter controlling sharpness
\end{itemize}

This objective can be interpreted as a $(K+1)$-way ranking problem where the model must identify the correct positive sample among negatives. For this task this means the aim is for a given protein to score conditions resulting in a crystal higher than those who are less likely to. 

A common instantiation is the \textit{instance discrimination} task, where each sample is treated as its own class. Positive pairs are generated via augmentations of the same instance, while negatives come from different instances. The model thus learns invariance to transformations while preserving instance-level distinctions. For a given protein, positive samples can be generated from experimentally validated conditions by applying small perturbations, such as slight changes in concentration or pH, whereas negative samples are drawn from the global pool of all other conditions.

Crucially, the choice of augmentations defines the invariances encoded in the learned representation. Strong augmentations prevent trivial solutions and encourage semantic feature learning. However, since proteins are extremely sensitive to slight changes in condition large perturbations can create false positives and hinder training. 
Often contrastive learning relies on defining a similarity distribution $p^+(x^+|x)$ and a dissimilarity distribution $p^-(x^-|x)$. 
Training proceeds by sampling mini-batches containing multiple positive and negative pairs. A common problem is representation collapse, where all inputs map to the same representation. This can be avoided using temperature scaling, diversity in batches.
Contrastive learning provides a general framework for learning representations by comparing samples rather than predicting labels. Its effectiveness stems from:
\begin{itemize}
    \item Flexible definition of similarity
    \item Strong inductive bias via augmentations
    \item Scalable training with large negative sets
\end{itemize}

Contrastive learning is well-suited for this task since the end goal is to effectively rank conditions for a given protein. 