\subsection{Condition Embedding}\label{sec:methods/feature_engineering/condition_embedding}

One approach to obtain a numerical representation of chemical compounds is tokenization, a technique commonly used in natural language processing, that converts strings into sequences of integer values representing discrete tokens. The simplest tokenization strategy assigns each chemical compound a unique identifier, effectively treating each compound as an independent token. However, chemical nomenclature and representations often contain meaningful substrings, such as prefixes (e.g., mono- or di-) and suffixes (e.g., -ate), which encode chemically relevant information. Consequently, a more expressive approach is substring tokenization, where compounds are decomposed into smaller chemical subunits that can be represented individually.

Despite these advantages, purely token-based representations require the model to infer chemical properties solely from contextual co-occurrence patterns of tokens within the dataset. This limits the extent to which the representation explicitly reflects underlying physicochemical characteristics. To embed crystallization conditions in a more physically meaningful manner, this thesis therefore adopts an additional representation strategy. Namely, a numerical feature representation derived from physicochemical descriptors.

The numerical feature representative was created using the following procedure. After parsing the condition from the free text a nomenclature was defined of valid chemicals. For all the chemicals in that dictionary the \gls{smiles} notation was defined. 
Introduced by \cite{Weininger1988}, \gls{smiles} is a text-based chemical notation designed for computational processing. It represents a molecule as a linear string of ASCII characters encoding its underlying molecular graph, including atoms, bond types, branching, ring closures, and formal charges. Thus, \gls{smiles} captures the connectivity and chemical structure of a compound in a machine-readable format.

For example, ammonium sulfate ($(NH_4)_2SO_4$) is represented as: $$[NH4+].[NH4+].[O-]S(=O)(=O)[O-]$$

Dots indicate non-covalently connected species, here two ammonium ions and one sulfate ion. Charged atoms are written in square brackets, such as `[NH4+]` for the ammonium ion. The sulfate ion is encoded by specifying sulfur as the central atom (`S`), double-bonded oxygens using `=O`, and negatively charged oxygens as `[O-]`.

Overall, \gls{smiles} provides a compact representation of molecular structure by describing the atoms and their connectivity rather than their spatial arrangement, making it well suited for computational modeling and machine learning applications.

The \gls{smiles} code for the chemicals was tracked down using \href{https://pubchem.ncbi.nlm.nih.gov/}{PubChem}. \href{https://pubchem.ncbi.nlm.nih.gov/}{PubChem} is a database of chemical information managed by the U.S. National Institutes of Health and captures structures, physical properties of small molecules. 

Using the \gls{smiles} representation, a multitude of molecular descriptors are extracted for each chemical. These aim to properly represent the physical properties of the chemicals, the features, all computed with RDKit, can be found in \autoref{tab:feature_table} in \autoref{app:methods}.

Additionally, each compound is also represented by a Morgan fingerprint derived from its \gls{smiles} representation. \citet{Morgan1965} defines circular substructure descriptors that capture local atom environments around each atom up to a given radius. By hashing these substructures into a fixed-length bit vector, they provide a continuous representation of molecular topology suitable for machine learning. This way, functional groups, aromatic rings, local polarity patterns, repeating structural motifs and substructure frequencies can be encoded. In this work, a count-based Morgan fingerprints of length 2048 and radius 2 is chosen. The hashing procedure in the traditional (binary) Morgan fingerprint only encodes whether a particular circular substructure is present in the molecule or not. In contrast, the count-based variant additionally records how often each substructure occurs. Instead of a binary indicator (0/1), each fingerprint dimension therefore contains the frequency of the corresponding hashed substructure, allowing repeated structural motifs to be represented more explicitly.
